---
title: "Cultural Domain Analysis and MDS"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
tutorial:
  id: "CDA"
  version: 0.5

---
```{r setup, include=FALSE}
library(AnthroTools)
library(learnr)
library(gradethis)
library(psych)
knitr::opts_chunk$set(exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(echo = TRUE)
learnr::tutorial_options(
  exercise.timelimit = 60)
```


## Useful links
- [Freelisting slides from Analytic Technologies](http://www.analytictech.com/mb870/slides/Freelisting.pdf)
- [Anthrotools](https://anthrotools.wordpress.com/)
- [Research Methods in Anthropology by Russell Bernard](https://books.google.com/books?id=2Fk7DwAAQBAJ&printsec=frontcover&source=gbs_ge_summary_r&cad=0#v=onepage&q&f=false)
- [Pile sort slides from Borgatti](http://www.analytictech.com/borgatti/etk3.htm)

<a class="btn btn-primary btn-sm" href="https://maddiebrown.github.io/ANTH630/ANTH630syllabus.html" role="button">ANTH630 Home</a>

## Getting started

This lesson will teach you how to work with and analyze cultural domain data using R. This lesson draws in part on datasets and formulas from the [Anthrotools package](https://anthrotools.wordpress.com/). You can get the data for this lesson [here](https://maddiebrown.github.io/ethnoecology/FruitList.csv) and [here](https://maddiebrown.github.io/ethnoecology/pilesort_sampledata.csv).

## Load packages and data

```{r results=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2) 
library(tidyr)
```

Second, let's load the data we plan to use and assign it a name. Here we will load the `FruitList` data. Take a look at the `head()` and `str()` of this spreadsheet. How do you think the data have been organized? What do the rows and columns represent?

```{r prepare-fruitlist}
FruitList <- read.csv("D:/BSOS_Anthropology_R_Tutorial/ANTHROPOLOGY_R_TUTORIALS/data/FruitList.csv")
head(FruitList)
str(FruitList)
```

We can see that we have 75 rows with 4 variables. The `X` variable is a numeric index autocreated by excel. The `Subj'`variable is a numeric identifier for each interview respondent. The `Order` variable represents the order in which each fruit was named by each respondent. The `CODE` variable contains each fruit named in the freelists.

### Try it

How many respondents and unique fruits ('items') are in the `FruitList` data? What are the unique fruits? Write down the necessary code below!

```{r exercise1,exercise.setup= "prepare-fruitlist", exercise=TRUE}
```

```{r exercise1-solution}
length(unique(FruitList$Subj))
length(unique(FruitList$CODE))
unique(FruitList$CODE)
```

```{r exercise1-check}
grade_code("Excellent!")
```

From these few operations we now know a bit more about the dataset. There are 20 individuals who collectively named a total of 8 unique fruits. 

Next, let's calculate the average number of fruits named per person and explore the distribution of list lengths across responses.

```{r}
FruitList <- read.csv("D:/BSOS_Anthropology_R_Tutorial/ANTHROPOLOGY_R_TUTORIALS/data/FruitList.csv")
# How many responses per subject
FruitList %>% group_by(Subj) %>% count()
# What is the average number of responses per subject?
FruitList %>% group_by(Subj) %>% count() %>% ungroup() %>% summarise(mean=mean(n))
```

```{r results='hide'}
# We can also do this in two parts by creating a new dataframe of the summarized data and then calculating the mean value in the "n" column
BySubject <- FruitList %>% group_by(Subj) %>% count()
mean(BySubject$n)
# This method also allows you to generate additional summary information
summary(BySubject$n)
# We can also make a table showing how many respondents had lists of length 2, 3, 4, and 5
table(BySubject$n)
```

## Item frequency analysis

Next, let's analyze how frequently each fruit was mentioned. For this analysis we are going to create a new dataframe object called 'ByFruit' that groups the data according to each unique fruit and counts the number of rows.

```{r results='hide'}
ByFruit <- FruitList %>% group_by(CODE) %>% count() %>% arrange(desc(n))
ByFruit
```

### Try it

Plot the number of times each type of fruit is mentioned by interviewees. Make a bar plot with flipped coordinates so that each fruit name is on the y axis and the number of mentions are on the x axis. 

```{r exercise2, exercise.setup="prepare-fruitlist", exercise=TRUE}

```

```{r exercise2-solution}
ggplot(ByFruit, aes(x=reorder(CODE,n),y=n)) + geom_bar(stat="identity") +coord_flip() +ggtitle("Frequency of fruit mentions") + labs(x="Fruit", y="n")
```

```{r exercise2-check}
grade_code("Excellent!")
```

### Scree Plots

Another way to visualize these data is through a scree plot. In these types of plots data are arranged such that each point represents a single item, the y-value for which is its frequency of mention in the dataset. These types of plots can be used to quickly identify trends and cut-off points in the data.

```{r}
ggplot(ByFruit, aes(x=reorder(CODE,desc(n)),y=n,group=1)) + geom_point() + geom_line()
```

### Frequency vs. rank

In freelists, we often expect the more highly ranked items to also show up more frequently. The code below plots the frequency of fruit mention against its average rank to explore this pattern.

```{r}
# First, we make a new 'ByFruit' object that includes the fruit frequency, top rank, and average rank.
ByFruit <- FruitList %>% group_by(CODE) %>% summarise(Frequency=n(),topRank=min(Order),avgRank=mean(Order))
ByFruit # Look at our new object
#Plot these varables
plot(ByFruit$Frequency,ByFruit$avgRank)
```

With more observations there will often be a trend where the top ranking items show up most frequently. This is a toy dataset with only a few samples, so this may explain why the trend doesn't seem supported, but try this out with your own freelist data.

## Salience calculations

One of the main insights than can be learned from freelist data is the relative cultural salience of items in particular domains. For example, in the domain of *household chores*, we could determine whether vacuuming, washing dishes, shoveling the driveway, or feeding the snake are considered more salient or central to the idea of *household chores* compared to other tasks. We might find variations based on individual attributes or the cultural context in which the question is asked. Salience often mirrors frequency, but the calculation is a bit more complicated as it considers both an item's frequency of mention and the order in which it is usually listed.

Luckily, the package `AnthroTools` has a built in salience calculation function that can do the math for us. The code below calculates the salience of each fruit listed in the context of each individual interviewee.

```{r new-fruitlist}
FruitListSalience <- CalculateSalience(FruitList,Order="Order",Subj="Subj",CODE="CODE")
# Note: I have included the arguments for Order, Subj and CODE for illustrative purposes. Because these column names match the arguments, you do not actually need to include them in this case. However, when you are working with your own datasets you may have different column names, so it is helpful to keep the underlying structure of functions in mind when you deploy them.
FruitListSalience
```
The above code calculates the salience for each item by respondent. If you inspect the results, you'll see that the first item in each list has a salience of 1, with each subsequent item decreasing in relative salience. This is useful for understanding how an individual thinks about the domain of fruits, but what if we are interested in knowing how salient **apples** are across all responses? We can calculate the salience of particular items as well with the SalienceByCode() function. 

### Try it

Try running the `SalienceByCode()` function on the new salience dataframe we made above.

```{r exercise3, exercise.setup="new-fruitlist", exercise=TRUE}
```

```{r exercise3-solution}
SalienceByFruit <- SalienceByCode(FruitListSalience,dealWithDoubles="MAX") 
# The dealwithdoubles argument tells R what to do if a respondent lists the same item twice. There are a few different options available for this, the right one to pick will depend on your data and research question.
  
```

```{r exercise3-check}
grade_code("Excellent!")
```

From this analysis we can see that there are slight differences in the mean salience and Smith's S. Smith's S considers the length of lists in its calculation of salience (More info [here](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3551534/). 

Now let's plot the Smith's S results in decreasing order and add a vertical line at the 0.1 mark. This range is generally considered a benchmark level for assessing item salience in freelists.

```{r}
SalienceByFruit <- SalienceByCode(FruitListSalience,dealWithDoubles="MAX") 
ggplot(SalienceByFruit, aes(x=reorder(CODE,SmithsS),y=SmithsS)) + geom_bar(stat="identity") +coord_flip() +ggtitle("Fruit Salience") + labs(x="Fruit", y="Smith's S") + geom_hline(yintercept=0.1)
```

From this plot, it looks like most of the fruits could be considered salient in this dataset. However, although pears were mentioned slightly more frequently than bananas and plums, their salience is lower when their overall order within lists and other factors are taken into account. 

## Comparing across groups

It is also possible to compare item salience across different groups of respondents. The following example comes from the AnthroTools package which includes a sample grouping of the FruitList data. First, let's load the new dataset. Because it is included in the AnthroTools package, we can load it directly with the data() function.

```{r prepare-worldlist}
data("WorldList")
WorldList
```

### Try it 

Try calculating the salience for each fruit, adding in an argument to differentiate responses by `GROUPING`. The argument is conveniently called GROUPING. First calculate the salience of each item by response, then calculate the salience of each item. Refer to the code we ran earlier in this lesson as a template to write your solution.

```{r exercise4, exercise.setup="prepare-worldlist", exercise=TRUE}
```

```{r exercise4-solution}
FL1 <- CalculateSalience(WorldList, Order="Order",Subj="Subj",CODE="CODE",GROUPING="GROUPING")
#Note, this function will not run without adding in the GROUPING argument.
FL2<- SalienceByCode(FL1,GROUPING="GROUPING",dealWithDoubles = "MAX")
```

```{r exercise4-check}
grade_code("Excellent!")
```

## Frequency by group 

In addition to evaluating salience, we might be interested in the percentage of respondents from each group who named a particular item. The code below creates a new object where the data are grouped according to the GROUPING variable.

```{r}
frequencybygroup <- WorldList %>% group_by(GROUPING) %>% mutate(GroupN=length(unique(Subj))) %>% ungroup %>% group_by(GROUPING, CODE, GroupN) %>% summarise(totalResponses=n(),nRespondents=length(unique(Subj)),percentRespondents=round(length(unique(Subj))/first(GroupN)*100,2)) %>% arrange(GROUPING, desc(percentRespondents))
```

We can then plot the results of this grouping with a facet wrap graph. 
```{r}
ggplot(frequencybygroup, aes(x=reorder(CODE,desc(percentRespondents)),percentRespondents)) + geom_bar(stat="identity") + coord_flip() +ggtitle("Frequency of fruit mentions by site") + labs(x="Fruit", y="n") + facet_wrap(vars(GROUPING))
```
It looks like in all cases, apples were mentioned more frequently, but the different sample populations differ in their rates of mentioning oranges and bananas. One thing you will notice in this graph is that although the fruits are listed in descending order for the "ISLAND" sample, the order is not meaningful for the other sites. This is because in the faceted plot, the y axis is the same for each site, making it easier to compare across samples. 

## Comparing freelist lengths across groups

We can also compare the lengths of lists across each group. This shows us that the average list lengths were not very different across the groups, but were slightly higher in the Island population.

```{r}
WorldList %>% group_by(GROUPING,Subj) %>% summarise(n=n()) %>% ungroup %>% group_by(GROUPING) %>% summarize(nResponse=n(),avgLength=mean(n),maxLength=max(n),minLength=min(n),medianLength=median(n))
```

## Advanced Freelist Analysis

### Co-occurence tables
Beyond investigating how many fruits are listed how frequently within the *fruit* domain, we are also interested in learning more about the structure of this domain. We can think of this as a type of mental map, how do people think about each item within the *fruit* domain relative to the other items? 

One way to investigate this question is through examining co-occurence of items within the same freelist. Essentially, how many times is each pair of fruits mentioned by the same respondent? We can create a co-occurence matrix with `tidyverse` tools or the `AnthroTools` package. 

### Try it
Using `FreeListTable()` make a table indicating whether or not each respondent mentioned a particular fruit. *Hint: This is a presence/absence table*.

```{r exercise5, exercise.setup="prepare-fruitlist", exercise=TRUE}
```

```{r exercise5-solution}
FruitListTable <- FreeListTable(FruitList, CODE = "CODE", Salience = "Salience", Subj = "Subj", 
    tableType = "PRESENCE")
```

```{r exercise5-check}
grade_code("Excellent!")
```

You can do the same operation with  `tidyverse`.

```{r}
#add new count column
FruitList$present <- rep(1)
# Spread into wide datatable. Note: spread function requires unique identifiers for rows, so here we remove any duplicate rows
FruitListWide <- FruitList %>% select(Subj,CODE,present) %>% unique %>% spread(CODE, present)
#convert NAs to 0
FruitListWide[is.na(FruitListWide)] <- 0
FruitListWide 
```

In this case, we have made a presence/absence matrix. That is, do the two items co-occur or not. You could also create a matrix that is weighted based on the number of co-occurences. the co-occurence matrix can now be used for clustering, MDS, network analysis and other analyses. 

## Multidimensional scaling

With the matrix we made above, we can analyze how similar or dissimilar each fruit it from one another using multidimensional scaling (MDS). The data manipulations can be a little tricky, but I've included the code below to get you started.

```{r}
# Here we use the FruitListWide dataframe we made above.
## Convert dataframe into presence/absence only by removing the first column. First we can assign the Subj column to the rownames
rownames(FruitListWide) <- FruitListWide$Subj
# Remove the subj column
FruitListWide<-FruitListWide[,-1]
# Now convert datasetbyword matrix to a wordbyword co-occurance matrix
FruitListWide <- as.matrix(FruitListWide)
FruitsBySubj <- t(FruitListWide)
# Look at data structure
FruitsBySubj
FruitsByFruits <- crossprod(FruitListWide)
## Make a distance matrix
FruitDistanceMatrix <- cmdscale(dist(FruitsBySubj))
FruitDistanceMatrixDF<- data.frame(FruitDistanceMatrix)
## Using binary mds method
FruitsBySubj %>% dist(method="binary") %>% cmdscale(eig=T, k=2) -> test2
test3 <- data.frame(test2$points) %>% #mds coordinates
  bind_cols(Fruits = rownames(FruitsBySubj)) %>% #bind sample names
  bind_cols(count=rowSums(FruitsBySubj)) #bind count by each fruit
# Look at results
test3
## Plot results
ggplot(test3,aes(x = X1,y = X2, label = Fruits)) +geom_text() + geom_point(aes(size=count),alpha=0.4,color="blue")+labs(size="Count",x="Dimension 1", y="Dimension 2") + ggtitle("MDS of FruitList")
```

This MDS shows the fruits by reponses, showing us how the fruits are mentioned by different types of respondents. We might also be interested in an MDS of the fruits by fruits co-occurence, which can be calculated through the code below.

```{r}
## Using binary mds method
FruitsByFruits %>% dist(method="binary") %>% cmdscale(eig=T, k=2) -> test4
test5 <- data.frame(test4$points) %>% #mds coordinates
  bind_cols(Fruits = rownames(FruitsByFruits))#bind sample names
# Look at results
#test5
## Plot results
ggplot(test5,aes(x = X1,y = X2, label = Fruits)) +geom_text(fontface="bold",size=2.5,alpha=0.6,position=position_jitter(width=0.05, height=0.005)) + geom_point(size=NA)+labs(x="Dimension 1", y="Dimension 2") + ggtitle("MDS of FruitbyFruit")
```

This plot shows us which fruits are more central to the domain "fruits" across all the respondents and which are more peripheral. The fruits in the center of the plot co-occur more frequently since they are in more of the freelists in the sample. Borgatti has a useful presentation explaining [more](http://www.analytictech.com/mb870/slides/Freelisting.pdf).


## Pile Sorts

### Data wrangling

In addition to freelist data, pile sorts are a common tool used in cultural domain analysis. I've made a sample dataset of pilesorts of 40 trees from 9 individuals. You can load the data in the same way we loaded the fruitlist data. We are doing a bit of data wrangling to get the data into the right format for MDS scaling. Hopefully this helps you understand a bit about how data can be shaped and reshaped no matter what form it comes to you in!

```{r prepare-pilesort, results='hide'}
treepilesort <- read.csv("D:/BSOS_Anthropology_R_Tutorial/ANTHROPOLOGY_R_TUTORIALS/data/pilesort_sampledata.csv")
rownames(treepilesort) <- treepilesort$Tree
# Remove tree column
treepilesort <- treepilesort[,-1]
# Flip so rows are individual responses
PersonbyTree <- t(treepilesort)
PersonbyTree<-as.data.frame(PersonbyTree) # make the output into a dataframe
```

### Try it

#### 1. Currently the row names correspond to each respondent code (letters from A to I). Append these rownames to the dataframe as a new column named "Person".

```{r exercise6, exercise.setup="prepare-pilesort", exercise=TRUE}
```

```{r exercise6-solution}
PersonbyTree$Person <- rownames(PersonbyTree)
```

```{r exercise6-check}
grade_code("Excellent!")
```

#### 2. Convert the dataframe from wide to long format, where each row conveys which respondent placed which fruit into which pile. Hint: if using `gather()` you might need a third argument `maple:coconut`. You might also consider using `pivot_longer()`.

```{r exercise7, exercise.setup="prepare-pilesort", exercise=TRUE}
```

```{r exercise7-solution}
PersonbyTreeLong <- PersonbyTree %>% gather(Species,Pile,maple:coconut)
```

```{r exercise7-check}
grade_code("Excellent!")
```

The above operation can be done using `tidyverse` as well!

```{r eval=FALSE}
#another way with tidyverse
PersonbyTree %>% pivot_longer(!Person, names_to = "Species", values_to = "Pile")
PersonbyTree %>% pivot_longer(-Tree,names_to="Species",values_to="Pile") #Another way to do this with new tidyverse
```


Now that our data are in a long format we can add in a new unique identifier for each person_pile combination. This step is required because the piles have numeric names that repeat across respondents. Next we can make a presence/absence column that will be useful for creating a new matrix based on whether or not a tree occurs in a particular pile. 

```{r}
# Add new column which is person ID and pile ID combined together
PersonbyTree$Person <- rownames(PersonbyTree)
PersonbyTreeLong <- PersonbyTree %>% gather(Species,Pile,maple:coconut)
PersonbyTreeLong<- PersonbyTreeLong %>%  mutate(PersonPile=paste(Person,Pile,sep="_"))
PersonbyTreeLong$present <- rep(1)
```

Now we can convert the data back into a wide format, where each column is a different tree species and the values represent presence or absence of the tree.

```{r}
# Spread into wide datatable. Note: spread function requires unique identifiers for rows, so here we remove any duplicate rows
TreeWide <- PersonbyTreeLong %>% select(PersonPile,Species,present) %>% unique %>% spread(Species, present)
# Convert NAs to 0
TreeWide[is.na(TreeWide)] <- 0
head(TreeWide)
```

This dataframe looks great. The multidimensional scaling functions in R take a matrix however, so we will convert the dataframe into two different matrices following the same procedure we used for the fruit data.

```{r}
## Convert dataframe into presenve/absence only by removing the first column. First we can assign the Subj column to the rownames
rownames(TreeWide) <- TreeWide$PersonPile
# Remove the subj column
TreeWide<-TreeWide[,-1]
# Now convert datasetbyword matrix to a wordbyword co-occurance matrix
TreeWide <- as.matrix(TreeWide)
TreeByPile <- t(TreeWide)
# Look at data structure
TreeByPile
```

## Multidimensional scaling

Now that our data are in the proper format, we can examine them with MDS. 

```{r}
#FIRST looking at tree by pile
##distance matrix
TreeByPileMatrix <- cmdscale(dist(TreeByPile))
TreeByPileMatrixDF<- data.frame(TreeByPileMatrix)
##using binary mds method
TreeByPile %>% dist(method="binary") %>% cmdscale(eig=T, k=2) -> test2
test3 <- data.frame(test2$points) %>% #mds coordinates
  bind_cols(Trees= rownames(TreeByPile)) %>% #bind sample names
  bind_cols(count=rowSums(TreeByPile)) #bind count by each fruit
# Look at results
test3

##plot results
ggplot(test3,aes(x = X1,y = X2, label = Trees))+geom_text(cex=3,alpha=0.5,position=position_jitter(width=0.05, height=0.05)) + geom_point(size=NA)+labs(x="Dimension 1", y="Dimension 2") + ggtitle("MDS of Trees by Pile")

```

This plot shows an MDS of the trees by which pile they are in. It shows us there are 4-5 different types of piles usually made by respondents. This can help in interpretation, but often we are most interested in the relationship between each tree species and the other tree species. For this, we work with a `TreebyTree` matrix.

```{r}
# Make a treebytree dataframe
TreebyTree <- crossprod(TreeWide)
head(TreebyTree) 
## Using binary mds method
TreebyTree %>% dist(method="binary") %>% cmdscale(eig=T, k=2) -> test4
test5 <- data.frame(test4$points) %>% #mds coordinates
  bind_cols(Trees = rownames(TreebyTree))#bind sample names
# Look at results
test5
```

Then we can plot the results.

```{r}
## Plot results
ggplot(test5,aes(x = X1,y = X2, label = Trees)) +geom_text(fontface="bold",size=2,alpha=0.4,position=position_jitter(width=0.05, height=0.02)) + geom_point(size=NA)+labs(x="Dimension 1", y="Dimension 2") + ggtitle("MDS of Trees")
```

This plot can be a bit difficult to read at times depending on how the labels are plotted. You can rerun the code until a more legible version of the graph is generated. You can also adjust the label transparency, size, and jitter to improve graph legibility. So far, it looks like the people in this sample consider coconuts and bananas to be very different from all the other trees, while pines exhibit their own cluster but are still more similar to other trees than to the coconut-banana cluster.

## Cluster analysis

In addition to multidimensional scaling, we can also examine the pilesort data through hierarchical clustering. This method allows us to understand the hiararchy of cluster similarities to one another. The code below allows us to create a cluster dendrogram of the tree data.

```{r}
# Then we can run a cluster analysis on the pile sort data
# Code adapted from: https://www.r-bloggers.com/clustering-music-genres-with-r/. Check it out for examples of how to take this analysis further.
#first turn data into matrix and remove diagonal
TreebyTreeMat <- as.matrix(TreebyTree)
diag(TreebyTreeMat) <- 0
#make distance matrix of co-occurence data
TreeDistMat <- dist(TreebyTreeMat)
#perform hierchical clustering on data
TreeHC <- hclust(TreeDistMat, method="ward.D")

plot(TreeHC,xlab="Tree species", sub="Ward's method")

```

Depending on which height the graph is cut at, different numbers of tree clusters are created. This visualization can help us better understand how similar or dissimilar groups of trees are to one another, depending on the level of similarity used as a benchmark. For example, if we cut the dendrogram at a height of 100, there will be two cluster, which roughly correspond to *conifers* and *non-conifers*. However, if we cut the dendrogram at a height of 20, we will have 5 clusters, which could be named something like *conifers*, *oaks*, *non-oak non-fruit/nut bearing*, *fruit trees*, and *nut trees*. The level at which you choose to define clusters depends on both the data analysis results and questions of your study.




