---
title: "ANTH630: Week 4 - Visualizing data"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
tutorial:
  id: "Week4VisualizingData"
  version: 0.5
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
knitr::opts_chunk$set(exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(echo = TRUE)
learnr::tutorial_options(
  exercise.timelimit = 30)
```

Tutorial written by Madeline Brown and modified for learnr

### Useful Links
* [R Graph Gallery](https://www.r-graph-gallery.com/)
* [R Graphics Cookbook](https://r-graphics.org/)

<a class="btn btn-primary btn-sm" href="https://maddiebrown.github.io/ANTH630/ANTH630syllabus.html" role="button">ANTH630 Home</a>


## Getting started with ggplot2

Unlike plotting with base R, `ggplot2` relies on adding different layers onto a plot one at a time to create a complete figure. These layers are called `geoms`, and include axes, labels, points, or other information. The aesthetics of `geoms` are further customized with the `aes()` argument. 

```{r}
library(ggplot2)
```

### Creating a plot in layers

In ggplot2, figures are created one layer at a time. Let's work with the `diamonds` dataset to explore ggplot2's functions.

```{r}
ggplot(diamonds, aes(carat,price))
```

This output gives us an empty plot. One thing to notice is that the axis limits are set for the data referenced in the `aes()` argument, though the points are not plotted.

Now we can add in points with the `geom_point()` function.
```{r}
ggplot(diamonds, aes(carat,price)) + geom_point()
```

You can also plot multiple layers of data on the same plot.
```{r}

ggplot(diamonds, aes(carat,price)) + geom_point() + geom_smooth()

```

Let's finish the plot with some descriptive labels.

```{r}

ggplot(diamonds, aes(carat,price)) + geom_point() + geom_smooth() + ggtitle("Diamond prices by carat") + labs(x="Carat", y="Price (USD)")

```

Now you've mastered building plots in layers.

### Customizing aesthetics 
There are multiple aesthetic parameters that can be customized in ggplots. This includes: color, fill, linetype, size, shape, font, and more. It just depends on which `geom` you are working with. We will explore some of these graphical parameters further as this tutorial introduces different geoms. [Here](https://ggplot2.tidyverse.org/articles/ggplot2-specs.html) is a vignette about aesthetic customization in ggplot2.


### Different geoms

```{r eval=FALSE}

geom_col()
geom_point()
geom_line()
geom_smooth()
geom_histogram()
geom_boxplot()
geom_text()
geom_density()
geom_errorbar()
geom_hline()
geom_abline()

```

## Bar plots

Bar plots are great for showing frequencies or proportions across different groups. Let's return to the otter dataset we analyzed last week. First, let's calculate the total number of otters observed per site and then plot this in a bargraph with `ggplot2`.
```{r}
otter <- read.csv("https://maddiebrown.github.io/ANTH630/data/sea_otter_counts_2017&2018_CLEANDATA.csv")
notterpersite <- aggregate(x=n_otter ~ site_name, FUN=sum, data=otter)
ggplot(notterpersite, aes(x=site_name, y=n_otter)) + geom_col()
```

### Try it
Hmm...there are way too many columns in the plot above to make it a meaningful barchart. Subset out the top 5 sites with the most otter sightings and recreate the above chart. 

*To get past this exercise, you'll need to submit the code exactly as it is in the solution with the same variable names.*
```{r prepare-otter}
otter <- read.csv("https://maddiebrown.github.io/ANTH630/data/sea_otter_counts_2017&2018_CLEANDATA.csv")
notterpersite <- aggregate(x=n_otter ~ site_name, FUN=sum, data=otter)
ggplot(notterpersite, aes(x=site_name, y=n_otter)) + geom_col()
```

```{r exercise1-1, exercise.setup="prepare-otter", exercise = TRUE}
# notterpersite <-
# top5 <-
```

```{r exercise1-1-solution}
notterpersite<- notterpersite[order(notterpersite$n_otter,decreasing=T),]
top5 <- notterpersite[1:5,]
ggplot(top5, aes(x=site_name, y=n_otter)) + geom_col()
```

```{r exercise1-1-check}
grade_code()
```


Alright, this figure is looking better. But suppose we want to reorder the categories on the X-axis such that they are in order from fewest to greatest number of otter sightings?

We can reorder the cateogries according to the value in the n_otter column. First, inspect the structure of the site_name data. You'll notice that the remaining categories are in alphabetical order. 

```{r, include=FALSE}
# this code is to include the notterpersite and top5 variables in the tutorial's R session 
# so the next code chunk has access to them. The exercises have their own R session
# so the other code chunk's won't be able to exercise the variables defined there.
notterpersite<- notterpersite[order(notterpersite$n_otter,decreasing=T),]
top5 <- notterpersite[1:5,]
```

```{r}
str(top5$site_name)
top5$site_name<-reorder(top5$site_name,top5$n_otter)
p1 <- ggplot(top5, aes(x=site_name, y=n_otter)) + geom_col()
p1
```


## Working with axes and labels

R will usually make default x and y limits, but sometimes we want to manually adjust these ranges. Let's adjust the ylimit of our otter plot to 500. In reality, we wouldn't want to adjust the axes in this case, but this represents the principle.

```{r}
p1 + ylim(0,500)
```

We can also flip the x and y coordinates. 
```{r}
p1 + coord_flip()

```

### Try it {data-allow-skip=TRUE}

Using the `breaks` and `labels` arguments in the `scale_y_continuous()` function, change the labels in the otter plot to any text of your choosing. Finish your plot with a `ggtitle()`

```{r prepare-p1, exercise.setup="prepare-otter", include=FALSE}
# chain exercise setup from previous exercise
notterpersite<- notterpersite[order(notterpersite$n_otter,decreasing=T),]
top5 <- notterpersite[1:5,]

top5$site_name<-reorder(top5$site_name,top5$n_otter)
p1 <- ggplot(top5, aes(x=site_name, y=n_otter)) + geom_col()
```

```{r exercise2-1, exercise.setup="prepare-p1", exercise = TRUE}

```

```{r exercise2-1-solution}
p1 + scale_y_continuous(breaks = c(50,100,150,200,250,300), labels=c("a few","some","more","a bunch","a lot","a ton")) + ggtitle("Otter Counts")
```

## Working with colors

With ggplot you can customize the colors of different components of the graph in numerous ways. Let's work with the otters graph we just made to see how ggplot understands color arguments.

### Try it {data-allow-skip=TRUE}

1. Using the fill and color arguments in geom_col, add colors to the otter plot.

```{r exercise3-1, exercise.setup="prepare-p1", exercise = TRUE}

```

```{r exercise3-1-solution}
ggplot(top5, aes(x=site_name, y=n_otter)) + geom_col(fill="firebrick", color="lightsalmon2")
```

2. What happens if you put the same arguments into the aes() argument of the ggplot() function?

```{r exercise3-2, exercise.setup="prepare-p1", exercise = TRUE}

```

```{r exercise3-2-solution}
ggplot(top5, aes(x=site_name, y=n_otter, fill="firebrick", color="lightsalmon2")) + geom_col()
```

3. What happens if you assign "fill" to the site name variable?

```{r exercise3-3, exercise.setup="prepare-p1", exercise = TRUE}

```

```{r exercise3-3-solution}
ggplot(top5, aes(x=site_name, y=n_otter, fill=site_name)) + geom_col()

ggplot(top5, aes(x=site_name, y=n_otter)) + geom_col(aes(fill=site_name))
```

There is a major difference between assigning color within the aes() argument or outside of it.

### Try it {data-allow-skip=TRUE}
**Stacked bar chart**

Suppose we wanted to take the top 5 otter sites and show the different observation counts for each year. Try using the fill argument to create this figure. You will also likely need to retransform the data while including the three columns of interest.

```{r exercise4-1, exercise.setup="prepare-otter", exercise = TRUE}

```

```{r exercise4-1-solution}
notterperyearsite <- aggregate(x=n_otter ~ site_name + year, FUN=sum, data=otter)
ggplot(notterperyearsite, aes(site_name,n_otter,fill=year)) + geom_col()
```

### Working with color palettes

What if we want to use color to represent a scale?

```{r}
# colors mapped to categorical variable

ggplot(diamonds, aes(carat, price, color=cut)) + geom_point()

# ggplot creates an automatic scale
ggplot(diamonds, aes(x, carat, color=price)) + geom_point()

# you can add a manual scale with two colors on either end
ggplot(diamonds, aes(x, carat, color=price)) + geom_point() + scale_colour_gradient(low = "lightpink", high = "darkmagenta")

# you can add a manual scale with two colors on either end and a clear midpoint
ggplot(diamonds, aes(x, carat, color=price)) + geom_point() + scale_colour_gradient2(low = "deeppink4", mid="white", high = "lightblue", midpoint=median(diamonds$price))

```

There are also a variety of built-in palettes in R that are useful. 


```{r}

# viridis is great for making color-blind friendly plots
ggplot(diamonds, aes(carat, price, color=price)) + geom_point() + scale_colour_viridis_c()

#color brewer has convenient built-in palettes and is usful for base R plotting
ggplot(diamonds, aes(carat, price, color=clarity)) + geom_point() + scale_colour_brewer(palette="Oranges")

ggplot(diamonds, aes(carat, price, color=clarity)) + geom_point() + scale_colour_brewer(palette="Set3")

# we can also make a custom rainbow palette, as we did with base r last week.
rpal<-rainbow(8)
ggplot(diamonds, aes(carat, price, color=rpal[clarity])) + geom_point() 


```

## Adding text
We can add text to a plot using the `annotate()` function. 

```{r}
ggplot(diamonds, aes(carat, price, color=rpal[clarity])) + geom_point() + annotate("text", x=4,y=5000,label="Diamond prices")

```

### Try it {data-allow-skip=TRUE}

Using the same diamonds plot as before. Make a new text annotation in the upper left corner that is large, bold, serif font, and a color other than black.

```{r prepare_rpal, include=FALSE}
rpal<-rainbow(8)
```

```{r exercise5-1, exercise.setup="prepare_rpal", exercise = TRUE}

```

```{r exercise5-1-solution}
ggplot(diamonds, aes(carat, price, color=rpal[clarity])) + geom_point() + annotate("text", x=1,y=15000,label="Diamond prices", family="serif",fontface="bold", size=6,color="purple")
```

### Adding text based on values in plot

We can also add text labels to points, bars, or other values in a plot. 

```{r}
ggplot(top5, aes(x=site_name, y=n_otter)) + geom_col(aes(fill=site_name)) + geom_text(aes(label=n_otter), vjust=1.5,color="bisque")
```

## Facet plots

Sometimes it is helpful to create multiple plots on the same axis, but grouped by another variable. This is called faceting.

```{r}
library(MASS)

ggplot(Cars93, aes(Horsepower, Price)) + geom_point() + facet_wrap(Cars93$Type)

```

## Customizing the legend

R will automatically produce a legend in ggplot2. Often we need to customize this legend further. 
```{r}
ggplot(Cars93, aes(Horsepower, Price, color=Type)) + geom_point()

```
You can change the order of items in the legend. 
```{r}
ggplot(Cars93, aes(Horsepower, Price, color=Type)) + geom_point() + scale_color_discrete(breaks=c("Van", "Compact","Large","Midsize","Small","Sporty"))

```

You can also hide titles or the legend as a whole. More information in the [Cookbook for R](http://www.cookbook-r.com/Graphs/Legends_(ggplot2)

```{r}
ggplot(Cars93, aes(Horsepower, Price, color=Type)) + geom_point() + guides(color=guide_legend(title=NULL))
ggplot(Cars93, aes(Horsepower, Price, color=Type)) + geom_point() + guides(color=FALSE)

```

## Putting multiple plots together

When plotting in base R, we can change the parameters of the graphics output to include multiple plots. This is accomplished with `par(mfrow=c(#r,#c))`. It is usually best practice to reset the parameters after changing them, as otherwise all subsequent plots will also follow those parameters. 

```{r}
par(mfrow=c(1,2))
boxplot(cars$speed, main="Car speeds")
boxplot(cars$speed, main="Car speeds")
abline(h=mean(cars$speed),col="red",lty=2)
abline(h=median(cars$speed),col="blue",lty=2)
par(mfrow=c(1,1))
```

### Putting multiple plots together with ggplot

`ggplot` doesn't work with resetting the parameters. We can use functions from `gridExtra` to put multiple different plots in the same output page. Here is the example from today's lecture slides, where two barcharts are plotted next to one another (don't copy the poor y-axis limits seen here!). 
```{r}
library(gridExtra)
#example adapted from field 2012, common idea in data visualization
samp <- data.frame(month=c("May","June"), beachvisitors=c(100,175) )
p1 <- ggplot(data=samp, aes(x=month, y=beachvisitors)) + geom_col() + ggtitle("Beach visitors per month")
p2<- ggplot(data=samp, aes(x=month, y=beachvisitors)) + geom_col() + ylim(0,700) + ggtitle("Beach visitors per month")
grid.arrange(p1,p2,nrow=1)
```

## Saving graphical output

After making a beautiful graph, you can save it in a couple of different ways. First, in RStudio you can save the plot directly from the plots tab. Otherwise, you can also use several functions for saving graphical output. 

```{r}
samp <- data.frame(month=c("May","June"), beachvisitors=c(100,175) )
p1 <- ggplot(data=samp, aes(x=month, y=beachvisitors)) + geom_col() + ggtitle("Beach visitors per month")

#save as a pdf
#pdf("beach.pdf")
#ggplot(data=samp, aes(x=month, y=beachvisitors)) + geom_col() + ggtitle("Beach visitors per month")
#dev.off()

#saves the last ggplot to a file in the working directory
#ggsave("beach2.pdf")

```

## Themes

```{r}

ggplot(diamonds, aes(carat, price)) +geom_point() + theme_bw()

ggplot(diamonds, aes(carat, price)) +geom_point() + theme_classic()

ggplot(diamonds, aes(carat, price)) +geom_point() + theme_minimal()


```

Even more themes with [`ggthemes`](https://github.com/jrnold/ggthemes). Here is a helpful [gallery](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/).

```{r}
library(ggthemes)

ggplot(diamonds, aes(carat, price)) +geom_point() + theme_wsj()

ggplot(diamonds, aes(carat, price)) +geom_point() + theme_fivethirtyeight()

ggplot(diamonds, aes(carat, price)) +geom_point() + theme_gdocs()
```

### Further practice

An interesting exploratory data analysis and graphing tutorial using median household incomes in Maryland. Linked [here](https://rpubs.com/frankhecker/499524)

A pretty bubble plot tutorial can be found [here](https://www.r-graph-gallery.com/320-the-basis-of-bubble-plot.html).


