---
title: "ANTH630: Week 4 - Visualizing data"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
tutorial:
  id: "Week4VisualizingData"
  version: 0.5
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
knitr::opts_chunk$set(exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(echo = TRUE)
learnr::tutorial_options(
  exercise.timelimit = 60)
```

Tutorial written by Madeline Brown and modified for learnr

### Useful Links
* [R Graph Gallery](https://www.r-graph-gallery.com/)
* [R Graphics Cookbook](https://r-graphics.org/)

<a class="btn btn-primary btn-sm" href="https://maddiebrown.github.io/ANTH630/ANTH630syllabus.html" role="button">ANTH630 Home</a>


## Getting started with ggplot2

Unlike plotting with base R, `ggplot2` relies on adding different layers onto a plot one at a time to create a complete figure. These layers are called `geoms`, and include axes, labels, points, or other information. The aesthetics of `geoms` are further customized with the `aes()` argument. 

```{r}
library(ggplot2)
```

### Creating a plot in layers

In ggplot2, figures are created one layer at a time. Let's work with the `diamonds` dataset to explore ggplot2's functions.

```{r}
ggplot(diamonds, aes(carat,price))
```

This output gives us an empty plot. One thing to notice is that the axis limits are set for the data referenced in the `aes()` argument, though the points are not plotted.

Now we can add in points with the `geom_point()` function.
```{r}
ggplot(diamonds, aes(carat,price)) + geom_point()
```

You can also plot multiple layers of data on the same plot.
```{r}

ggplot(diamonds, aes(carat,price)) + geom_point() + geom_smooth()

```

Let's finish the plot with some descriptive labels.

```{r}

ggplot(diamonds, aes(carat,price)) + geom_point() + geom_smooth() + ggtitle("Diamond prices by carat") + labs(x="Carat", y="Price (USD)")

```

Now you've mastered building plots in layers.

### Customizing aesthetics 
There are multiple aesthetic parameters that can be customized in ggplots. This includes: color, fill, linetype, size, shape, font, and more. It just depends on which `geom` you are working with. We will explore some of these graphical parameters further as this tutorial introduces different geoms. [Here](https://ggplot2.tidyverse.org/articles/ggplot2-specs.html) is a vignette about aesthetic customization in ggplot2.


### Different geoms

```{r eval=FALSE}

geom_col()
geom_point()
geom_line()
geom_smooth()
geom_histogram()
geom_boxplot()
geom_text()
geom_density()
geom_errorbar()
geom_hline()
geom_abline()

```

## Bar plots

Bar plots are great for showing frequencies or proportions across different groups. Let's return to the otter dataset we analyzed last week. First, let's calculate the total number of otters observed per site and then plot this in a bargraph with `ggplot2`.
```{r}
otter <- read.csv("https://maddiebrown.github.io/ANTH630/data/sea_otter_counts_2017&2018_CLEANDATA.csv")
notterpersite <- aggregate(x=n_otter ~ site_name, FUN=sum, data=otter)
ggplot(notterpersite, aes(x=site_name, y=n_otter)) + geom_col()
```

### Try it
Hmm...there are way too many columns in the plot above to make it a meaningful barchart. Subset out the top 5 sites with the most otter sightings and recreate the above chart. 

<details>
<summary><span style="color:tomato">**Click for solution**</span></summary>
```{r}
notterpersite 
notterpersite<- notterpersite[order(notterpersite$n_otter,decreasing=T),]
top5 <- notterpersite[1:5,]
ggplot(top5, aes(x=site_name, y=n_otter)) + geom_col()
```
</details>

```{r exercise1-1, exercise = TRUE}

```

```{r exercise1-1-solution}
notterpersite 
notterpersite<- notterpersite[order(notterpersite$n_otter,decreasing=T),]
top5 <- notterpersite[1:5,]
ggplot(top5, aes(x=site_name, y=n_otter)) + geom_col()
```

```{r exercise1-1-check}
notterpersite 
notterpersite<- notterpersite[order(notterpersite$n_otter,decreasing=T),]
top5 <- notterpersite[1:5,]

grade_result(
  pass_if(~ identical(.result,ggplot(top5, aes(x=site_name, y=n_otter)) + geom_col()), "Good Job!"),
  fail_if(~ TRUE)
)
```
