---
title: "Getting Started with R Markdown"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
tutorial:
  id: "rmd"
  version: 0.5
---
```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(dplyr)
knitr::opts_chunk$set(exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(echo = TRUE)
learnr::tutorial_options(
  exercise.timelimit = 60)
```

<a class="btn btn-primary btn-sm" href="https://maddiebrown.github.io/ANTH630/ANTH630syllabus.html" role="button">ANTH630 Home</a>

## Intro 

With R Markdown you can typeset documents and code snippets. Organizing your problem sets into R Notebooks can be a clear way to present your process and results. This tutorial will help you get started. Download this code sourcefile to see how this document is made.

## Headings

###

Basic formatting is done following markdown conventions. 

`# Heading 1`

# Heading 1

`### Heading 2`

### Heading 2

`#### Heading 3`

#### Heading 3

`##### Heading 4`

##### Heading 4

You can also add inset boxes and other blocks. Insets are added with `> text`. Boxes are added around blocks of text that are tabbed twice.

> Inset

    Box

## Text Style

###

You can also adjust text style.

`**bold**`

**bold**

`*italic*`

*italic*

## Links and buttons

###

Links are added using square brackets followed by parentheses. 

`[UMD Anthro](https://anth.umd.edu/) `

[UMD Anthro](https://anth.umd.edu/) 

Buttons can be added using the following code:
`<a class="btn btn-primary btn-sm" href="https://anth.umd.edu/" role="button">UMD Anthro</a>`

<a class="btn btn-primary btn-sm" href="https://anth.umd.edu/" role="button">UMD Anthro</a>

## Lists

###

There are several ways to add lists. They can be customized with bullets, numbers, and nesting.


` - list item`

` - list item`

- list item

- list item

`* list item`

  `+ nested list item`
    
`* list item`

* list item

    + nested list item
    
* list item

`1. list item`

`2. list item`

`3. list item`

1. list item
2. list item
3. list item

## R code chunks

###

R code chunks are initiated with ` ```{r} ` and ended with ` ``` `. Any code in between these strings will be executed as normal R code.

```{r}
mean(2,5,7,3,8,8,8,4,2)
```

```{r}
library(ggplot2)
ggplot(cars, aes(speed,dist)) + geom_point(color="red", size=3)
```

```{r eval=FALSE}

interviews %>%
    filter(!is.na(memb_assoc)) %>%
    group_by(village, memb_assoc) %>%
    summarize(mean_no_membrs = mean(no_membrs)) %>%
  knitr::kable(caption = "We can also add a caption.", 
               col.names = c("Village", "Member Association", 
                             "Mean Number of Members"))

starwars %>% group_by(species) %>% summarise(num_char=n()) %>% arrange(num_char) %>% knitr::kable(caption="Star wars character counts", col.names=c("Species","# Characters"))
```


## Next steps

###

There are numerous themes that can be applied to R Markdown documents. [Themes](https://bootswatch.com/) (also [these](https://www.datadreaming.org/post/r-markdown-theme-gallery/)) vary depending on the specific type of markdown document.

In addition to R notebooks, markdown can also be used to make slides, websites, books, reports and more. [Here](https://rmarkdown.rstudio.com/gallery.html) is a gallery of R markdown products.

## Additional resources

###

* [R Markdown: The Definitive Guide. By: Yihui Xie, J. J. Allaire, Garrett Grolemund](https://bookdown.org/yihui/rmarkdown/)
* [Get Started with R Markdown](https://rmarkdown.rstudio.com/lesson-1.html)
* [R cheatsheets](https://www.rstudio.com/resources/cheatsheets/) (Includes R markdown sheet)
* [Interactive markdown tutorial](https://commonmark.org/help/tutorial/index.html)
* [R Markdown Cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/)
